\LaTeXe\ guide for the Journal of Functional Programming

### 2019-04-18 papers/jfp/4-schedulers.tex
Rework the intro a bit
Rework the schedulers chapter

\section{Build Systems \`a la Carte Schedulers}\label{sec-build}

\todo{Elaborate (it's very terse right now) with examples}

@@

\vspace{1mm}
We have emboldened two different aspects above: the part of the
build system responsible for scheduling tasks in the dependency order
(a `scheduler') can be cleanly separated from the part responsible for deciding
whether a key needs to be rebuilt (a `rebuilder'). We tackle each
aspect separately in subsections~\S\ref{sec-dependency-orderings}
and~\S\ref{sec-out-of-date}.

\subsection{The Scheduler: Respecting the Dependency Order}
\label{sec-dependency-orderings}

@@

\vspace{-2mm}
\subsubsection{Topological}\label{sec-topological}

The topological scheduler pre-computes a linear order of tasks, which when followed,
ensures the build result is correct regardless of the initial store. Given a
task description and the output \hs{key}, you can compute the linear order by
first finding the (acyclic) graph of the \hs{key}'s reachable dependencies, and
then computing a topological sort. However, as we have seen in~\S\ref{sec-deps},
we can only extract dependencies from an applicative task, which requires the
build system to choose \hs{c}~\hs{=}~\hs{Applicative}, ruling out dynamic
dependencies.

\vspace{-2mm}
\subsubsection{Restarting}\label{sec-restarting}

To handle dynamic dependencies we can use the following approach: build tasks in
an arbitrary initial order, discovering their dependencies on the fly; whenever
a task calls \hs{fetch} on an out-of-date key \hs{dep}, abort the task, and
switch to building the dependency \hs{dep}; eventually the previously aborted
task is restarted and makes further progress thanks to \hs{dep} now being up to
date. This approach requires a way to abort tasks that have failed due to
out-of-date dependencies. It is also not minimal in the sense that a task may
start, do some meaningful work, and then abort.

To reduce the number of aborts (often to zero) \Excel records the
discovered task order in its \emph{calc chain}, and uses it as the
starting point for the next build (\S\ref{sec-background-excel}).
\Bazel's restarting scheduler does not store the discovered order
between build runs; instead, it stores the most recent task dependency
information. Since this information may become outdated, \Bazel may
also need to abort a task if a newly discovered dependency is out of date.

\vspace{-2mm}
\subsubsection{Suspending}\label{sec-suspending}

An alternative approach, utilised by the \hs{busy} build system
(\S\ref{sec-general-build}) and \Shake, is to simply build dependencies when
they are requested, suspending the currently running task. By combining that
with tracking the keys that have already been built, one can obtain a minimal
build system with dynamic dependencies.

This approach requires that a task may be started and then suspended until
another task is complete. Suspending can be done with cheap green threads and
blocking (the original approach of \Shake) or using continuation-passing style
\cite{claessen_continuations} (what \Shake currently does).

>>>>

\section{Build System Schedulers}\label{sec-scheduler}

@@

We have emboldened two different aspects above: the part of the
build system responsible for scheduling tasks in the dependency order
(a `scheduler') can be cleanly separated from the part responsible for deciding
whether a key needs to be rebuilt (a `rebuilder'). In this section we discuss schedulers,
leaving rebuilders for \S\ref{sec-rebuilder}.

@@

\subsection{Topological Scheduler}\label{sec-topological}

The topological scheduler pre-computes a linear order of tasks, which when followed ensures
dependencies are satisifed, then executes the required tasks in that order. Computing such a linear
order is straightforward -- given a task description and output \hs{key} first find the (acyclic)
graph of the \hs{key}'s dependencies, then compute a topological sort. Taking the example from
Figure~\ref{fig-make}, we might compute the order:

\begin{enumerate}
\item \hs{util.o}
\item \hs{main.o}
\item \hs{main.exe}
\end{enumerate}

Given the dependencies, we could have equally chosen to build \hs{main.o} first, but \hs{main.exe} \emph{must}
come last.

The advantage of this scheme is simplicity -- compute an order, then execute tasks in that order.
In addition any missing nodes or cycles can be detected from the graph, and reported to the user
before any work has commenced.

The downside of this approach is that it requires the dependencies of each task in advance.
As we saw in~\S\ref{sec-deps}, we can only extract dependencies from an applicative task, which requires the
build system to choose \hs{c}~\hs{=}~\hs{Applicative}, ruling out dynamic dependencies.

\subsection{Restarting Scheduler}\label{sec-restarting}

To handle dynamic dependencies we cannot precompute a static order -- we must
interleave running tasks and ordering tasks. One approach is just to build tasks
in an arbitrary order, but if a task calls \hs{fetch} on an out-of-date key \hs{dep}
abort the task and build \hs{dep} instead. Returning to the example from
Figure~\ref{fig-make}, we might build in the order:

\begin{enumerate}
\item \hs{main.exe} (abort because it requires \hs{util.o})
\item \hs{util.o}
\item \hs{main.o}
\item \hs{main.exe}
\end{enumerate}

Here we start with \hs{main.exe} (chosen randomly), but discover it requires \hs{util.o}, so
instead start building \hs{util.o}. Next we choose to build \hs{main.o} (again, randomly),
before finally returning to \hs{main.exe} which now has all its dependencies available and
completes without error.

The approach works, but has a number of disadvantages. Firstly, it requires a technical mechanism
to abort a task -- easy enough in our theoretical setting of \hs{Task} but leads to engineering concerns
in the real world. Secondly, it is not minimal in the sense that a task may
start, do some meaningful work, and then abort, repeating that same work when restarted.

As a refinement, to reduce the number of aborts (often to zero) \Excel records the
discovered task order in its \emph{calc chain}, and uses it as the
starting point for the next build (\S\ref{sec-background-excel}).
\Bazel's restarting scheduler does not store the discovered order
between build runs; instead, it stores the most recent task dependency
information from which it can computed a linear order. Since this information may become outdated, \Bazel may
also need to abort a task if a newly discovered dependency is out of date.

\subsection{Suspending Scheduler}\label{sec-suspending}

An alternative approach, utilised by the \hs{busy} build system
(\S\ref{sec-general-build}) and \Shake, is to simply build dependencies when
they are requested, suspending the currently running task. Using
Figure~\ref{fig-make}, we would build:

\begin{enumerate}
\item \hs{main.exe} (suspended)
\begin{itemize}
\item \hs{util.o}
\end{itemize}
\item \hs{main.exe} (resumsed then suspended)
\begin{itemize}
\item \hs{main.o}
\end{itemize}
\item \hs{main.exe} (completed)
\end{enumerate}

As we see, the \hs{main.exe} is started first as it is the required target. It soon discovers
a dependency on \hs{util.o} and suspends to build \hs{util.o}, resumes a build before building \hs{main.o}
then continues its work to finish \hs{main.exe}.

This scheduler (when combined with a suitable rebuilder) provides a minimal build system
which supports dynamic dependencies.

The difficulties are all engineering in nature. To implement suspension there are two
standard approaches:

\begin{itemize}
\item Blocking threads or processes. This approach is relatively easy, but can require
significant resources, especially if a large number of tasks are suspended. In languages with
cheap green threads (e.g. Haskell) the approach is more feasible, and it was the original approach
taken by \Shake.
\item Continuation-passing style \cite{claessen_continuations} can allow the remainder of
a task to be captured, paused, and resumed later. Continuation passing is efficient, but requires
the build script to be architected to allow capturing continuations. \Shake currently uses
continuations.
\end{itemize}

While a suspending scheduler is theoretically optimal, it is only better in practice
than a restarting scheduler if the cost of duplicate work avoided outweighs the cost of suspending tasks.
